Reasoning Graphically about Quantum Computation

### 2008-03-06 papers/reasoning-graphically/paper.tex
work
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1403 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

Coecke and Duncan \cite{Coecke:2008jo} propose a formal algebraic
system for quantum computation built from the following collection of
generators:  
\begin{description}
\item[Objects:] the self dual object $Q$;
\item[Arrows:] two families of arrows $X$ and $Z$:
  \begin{align*}
  &\epsilon_Z : Q \to I, &\qquad&  \epsilon_X : Q \to I,\\
  &\delta_Z : Q \to Q \otimes Q, && \delta_X : Q \to Q \otimes Q, \\
  &\alpha_Z : Q \to Q &&  \alpha_X : Q \to Q   
  \end{align*}
  where $\alpha \in [0,2\pi)$, and in addition $H:Q\to Q$.
\end{description}
To each each arrow $f : A\to B$ we assign a formal adjoint $f^\dag : B
\to A$.  Each arrow is represented as a small graph; its adjoint is
the same graph written upside down.  We use colours to denote two families:
\begin{gather*}
  \epsilon_Z = \inlinegraphic{1.5em}{epsilon} \qquad
  \delta_Z = \inlinegraphic{1.5em}{delta} \qquad
  \epsilon_Z^\dag = \inlinegraphic{1.5em}{epsilondag} \qquad
  \delta_Z^\dag = \inlinegraphic{1.5em}{deltadag} \qquad
  \alpha_Z = \inlinegraphic{1.5em}{greenalpha} \qquad
\\  
  \epsilon_X = \inlinegraphic{1.5em}{redepsilon} \qquad
  \delta_X = \inlinegraphic{1.5em}{reddelta} \qquad
  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
  \alpha_X = \inlinegraphic{1.5em}{redalpha} \qquad
\end{gather*}
The $H$ arrow is denoted by \inlinegraphic{1.5em}{H}.  (The adjoints
for $H$, $\alpha_X$ and $\alpha_Z$ will be defined equationally.)  The
free compact closed category is then given by all graphs formed by
composing and tensoring these basic graphs.  Note  that since $Q$ is
self dual, that is $Q = Q^*$, we use \emph{undirected} graphs.

@@

\begin{figure}[t]
$$\begin{array}{ccc}
\inlinegraphic{8em}{spider_lhs_patt} & \;=\; & \inlinegraphic{6em}{spider_rhs_patt}
\end{array}$$
\label{fig:spider-thm-patt}\caption{The Spider Theorem expressed using graph patterns.}
\end{figure}

@@

*** add matching algo outline ***

@@

\subsection{Lifting}

*** lifting concrete laws to graph-pattern laws ***

\subsection{Adequacy}

An equation on graph patterns corresponds to
a possibly infinite set of equations betweens concrete graphs. We
might expect that such equations express too much: they may allow
rewrites which are not intended. We prove \emph{adequacy}, which
states that this is not the case. In particular, given an equation
between concrete graphs, $G = H$, the equations between the lifting of
these to variable node graphs by replacing boundary nodes with
variable nodes is a conservative extension. The proof follows directly
from the obervation that every instance of the lifted variable node
equation corresponds to compsition of $G$ and $H$ with some other
graph (the non-embedded part).

@@

\begin{center}
%% not needed, can be derived by refl + subst
%\prooftree
%A = B \in \Gamma
%\justifies
%\Gamma \vdash A = B
%\using\mbox{trivial}
%\endprooftree
%\quad\quad
\prooftree
\justifies
\Gamma \vdash A = A
\using\mbox{refl}
\endprooftree 
\quad\quad
\prooftree
\Gamma \vdash A = B
\justifies
\Gamma \vdash B = A
\using\mbox{sym}
\endprooftree
\quad\quad
\prooftree
\Gamma \vdash A = B \quad\quad
\Gamma \vdash C = D
\justifies
\Gamma \vdash (C = D[A/B])
\using\mbox{subst}
\endprooftree
\end{center}

\noindent where $\Gamma$ is the set of object-level axioms. For the
reflexivity rule, $A$ must be a well-formed pattern graph.

A new graph can be introduced using the reflexivity law. This can be
the left hand side of a conjecture, or a graph that needs to be
computed. By applying the \emph{subst} rule, intermediate results are
derived which can themselves be used to rewrite other rules and
conjetcures. In this way, the system allows derived rules to provide
an abbreviation for a combination of steps.

@@

% One of the main applications of

@@

\section{A Case Study in Quantum Computation}
\label{sec:case-study}
%The model of qantum computation in which we are interested is finite
%dimensional hilbert spaces.

\subsection{Node Expressions}
\label{ssec:node-expressions}

>>>>

Coecke and Duncan \cite{Coecke:2008jo} propose a formal algebraic
system for quantum computation built from the following collection of
generators:  
\begin{description}
\item[Objects:] are written $Q$ and are self dual;
\item[Arrows:] there are families of arrows $X$ and $Z$:
  \begin{align*}
  &\epsilon_Z : Q \to I, &\qquad&  \epsilon_X : Q \to I,\\
  &\delta_Z : Q \to Q \otimes Q, && \delta_X : Q \to Q \otimes Q, \\
  &\alpha_Z : Q \to Q &&  \alpha_X : Q \to Q   
  \end{align*}
  where $\alpha \in [0,2\pi)$, and in addition $H:Q\to Q$.
\end{description}
To each each arrow $f : A\to B$ we assign a formal adjoint $f^\dag : B
\to A$.  Each arrow is represented as a small graph; its adjoint is
the same graph written upside down.  We use colours to denote two families:
\begin{gather*}
  \epsilon_Z = \inlinegraphic{1.5em}{epsilon} \qquad
  \delta_Z = \inlinegraphic{1.5em}{delta} \qquad
  \epsilon_Z^\dag = \inlinegraphic{1.5em}{epsilondag} \qquad
  \delta_Z^\dag = \inlinegraphic{1.5em}{deltadag} \qquad
  \alpha_Z = \inlinegraphic{1.5em}{greenalpha} \qquad
\\  
  \epsilon_X = \inlinegraphic{1.5em}{redepsilon} \qquad
  \delta_X = \inlinegraphic{1.5em}{reddelta} \qquad
  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
  \alpha_X = \inlinegraphic{1.5em}{redalpha} \qquad
\end{gather*}
The $H$ arrow is denoted by \inlinegraphic{1.5em}{H}.  (The adjoints
for $H$, $\alpha_X$ and $\alpha_Z$ will be defined equationally.)  The
free compact closed category is then given by all graphs formed by
composing and tensoring these basic graphs.  Note  that since $Q$ is
self dual, that is $Q = Q^*$, we use \emph{undirected} graphs.

@@

\begin{figure}[t]
$$\begin{array}{ccc}
\inlinegraphic{12em}{spider_lhs_patt} & \;=\; & \inlinegraphic{9em}{spider_rhs_patt}
\end{array}$$
\label{fig:spider-thm-patt}\caption{The Spider Theorem expressed using
  graph patterns. The !-boxes are named $i$ and $j$. The variable
  nodes are white and named $a$ and $b$. The internal node's data,
  which is an angle, is written inside the node.}
\end{figure}

@@

Fortunately, it is quite easy to provide a decidable matching. This is
because the size of the unfolding that needs to be considered can be
bounded. The key observation is that a graph $G_1$ will never match a
graph with less non-variable nodes. Thus unfolding of $G_1$ can be
bounded by the number of non-variable nodes in $G_2$. While this gives
a generate and test style algorithm, it is not efficient. The
intuition for an efficient algorithm is to search through one graph
incrementally increasing the matched part and avoids symmetries in
matching. The development of an efficient algotihm is ongoing work.

%non-variable nodes than another graph $G_2$
%
%can algorithm as
%
%*** add matching algo outline ***

@@

\subsection{Lifting Axioms and Adequacy}

The axioms of an object system typically come from the semantics of
the unerlying system. For instance, the equations given in
Figure~\ref{fig:graph-quant-eqns} provide a system of reasoning about
quantum computation. In some cases, such equations can be provdied by
the underltying semantics. When such rules are expressed as graph
patterns, we replace the concrete representations boundary nodes with
variable nodes. This operation is called \emph{lifting}. An equation
on graph patterns corresponds to a possibly infinite set of equations
betweens concrete graphs. Thus we might worry that the lifted
equations express too much: they may allow rewrites which are not
intended.

We call the property that the lifted representation is a conservative
extension of the initial theory \emph{adequacy}. For models of compact
closed categories, the proof of adequacy is quite simple: given an
equation between concrete graphs, $G = H$, we observe that every
instance of the lifted equation corresponds to the original equations
composed with some graph. The graph is given by the unmapped subgraph
using the open embedding from the lifted equations onto their
instance. Thus if $G = H$ is true, then so is every instance of it's
lifting, and thus the lifting produces an adequate representation.

@@

\begin{center}
%% not needed, can be derived by refl + subst
%\prooftree
%A = B \in \Gamma
%\justifies
%\Gamma \vdash A = B
%\using\mbox{trivial}
%\endprooftree
%\quad\quad
\prooftree
%\mathit{valid}(A)
\justifies
\Gamma \vdash A = A
\using\mbox{refl}
\endprooftree 
\quad\quad
\prooftree
\Gamma \vdash A = B
\justifies
\Gamma \vdash B = A
\using\mbox{sym}
\endprooftree
\quad\quad
\prooftree
\Gamma \vdash A = B \quad\quad
\Gamma \vdash C = D
\justifies
\Gamma \vdash (C = D[A/B])
\using\mbox{subst}
\endprooftree
\end{center}

\noindent where $\Gamma$ is the set of object-level axioms and for the
reflexivity rule, $A$ must be a well-formed pattern graph. We assume
that axioms in $\Gamma$ meet the validity conditions described
earlier. These rules all preserve the validity conditions on equations
and thus the system as a whole ensures only valid rules are derived.

A new graph can be introduced using the reflexivity law (\emph{refl}).
By applying the \emph{subst} rule, intermediate results are derived
which can themselves be used to rewrite other rules and conjetcures.
In this way, the system allows derived rules to provide an
abbreviation for a combination of steps. Sets of rules can be applied
automatically to simulate computation in an the object domain. In
\S\ref{sec:case-study} we illustrate this for quantum computation.

@@

@@

\section{A Case Study in Quantum Computation}
\label{sec:case-study}

The model of quantum computation introduced in
\S\ref{sec:quotients-rewriting} provides an object formalism for
logical system. The equations from Figure~\ref{fig:graph-quant-eqns}
are lifted to graph patterns and the spider rule is encoded as shown
in Figure~\ref{fig:spider-thm-patt}. Although a complete graph-based
account of the underlying quantum model is ongoing work, these rules
provide an effective and useful initial set.

As well as these equations, recall that the object level formalism
must define the data held at the nodes and edges as well as provide a
suitable macthing algorithm for these. Our model of quantum
computation provides no data for the edges. The nodes, on the other
hand, are either a $H$ with no additional data (a Hadamard gate), or
have an \emph{angle} and are either the $X$ or $Z$ transforms. For
their part, angles are expressed as rational numbers which correspond
to the coefficient of $\pi$ in the corresponding matrix.

The result is that we give the $X$ and $Z$ nodes an \emph{angle
  expression}. When a node is within a !-box, this is a variable which
gets instantiated to a new variable in each of the unfoldings of the
!-box. When a node is not within a !-box, angle-expressions are a
mapping from a set of angle-variables to the corresponding rational
cooefficient. When an angle-expression contains a variable within a
!-box, this is interpreted as a sum of the variables that result from
it's unfolding. This rather simple expression language has a simple
normal form by lexographically ordering the angle-variable. Matching
then results in variables being instantiated and all affected nodes
normalising the result of the instantiation. The substitution rule
also has to make sure that a angle-variables in the rule being applied
are distinct from those in the expression being rewritten.

%The model of qantum computation in which we are interested is finite
%dimensional hilbert spaces.

%\subsection{Node Expressions}
%\label{ssec:cs:node-expressions}

\subsection{Computing the Quantum Fourier Transform}
\label{ssec:cs:quantum-transform}

The quantum fourier transform is among the most important quantum
algorithms, forming an essential part of Shor's
algorithm~\cite{Shor:PolyTimeFact:1997}.

\begin{figure}[t]
$$\begin{array}{cccc}
  & \inlinegraphic{2cm}{qft2} & = & \inlinegraphic{2cm}{qft3} \\ 
 & & & \\
= & \inlinegraphic{2cm}{qft6} & = & \inlinegraphic{2cm}{qft10}
\end{array}$$
\label{fig:quantum-transform}\caption{An example computation of the
  Quantum Fourier Transform, performed symbollically by rewriting.}
\end{figure}
